---
title: "On the curvature of the Phillips curve"
author: "Elisa Colson & Adriel Mazère"

format: 
  wordcount-html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    count-code-blocks: false

  wordcount-pdf:
    code: false
    count-code-blocks: false

execute:
  echo: false

editor: 
  markdown: 
    wrap: 72
    
prefer-html: true
---

# Setup

```{r, message = FALSE, warning = FALSE}
# Basic setup
here::i_am("Projet-Data-Management-Elisa-Adriel.Rproj")

# Libraries
library(vroom)
library(dplyr)
library(knitr)
library(tidyr)
library(lmtest)
library(sandwich)
library(ggplot2)
library(ggforce)
library(purrr)
library(broom)
library(here)
library(kableExtra)
```

::: callout-tip
## Link to the Github

You can find the link to the Github
[here](https://github.com/AdriMaze/Projet-Data-Management-Elisa-Adriel.git)
:::

# Motivation for the project

The dynamics in the labor market are the most complicated out of all
markets, because it mixes the humane reasoning and the economic one. The
Phillips curve has always been a reference tool because of the
simplicity of the relationship between prices and unemployment it
created. This relationship could not only provide an analytically point
of view about the conjuncture, but also a policy tool. In every
economics class, the Phillips curve was said to be negatively slopped,
but as one progresses, one realizes that this is less and less the case
and that multiple other underlying factors actually drive these two
forces. This is what we try to explore today.

# Research question

Building on this motivation, this project seeks to formalize the
relationship between inflation and unemployment and to examine how its
empirical properties have evolved across countries and over time.

The Phillips curve is a fundamental concept in economics, and the basis
of most neo-models, whether it be from classic or more keynesian
schools. It was first observed a negative relationship between
unemployment and inflation rate, but as time passes, and accross
countries as well, this relationship is less and less true. In many
advanced economies, periods of low unemployment have not been
accompanied by significant inflationary pressures, while inflation has
sometimes remained subdued despite tight labor market conditions.

However, empirical evidence accumulated over time suggests that this
relationship varies significantly across countries and historical
periods. Rather than reflecting a simple and stable trade-off, the
inflation–unemployment relationship appears to be influenced by a
broader set of economic, institutional, and policy-related factors. This
observation motivates an analysis that goes beyond the bivariate
relationship between inflation and unemployment, and instead considers
how changes in the surrounding economic environment may affect the
strength of this relationship.

::: text-center
**What drove the variations in the curvature of the Phillips curve since
the 1960s? In other words, how has the slope of the Phillips curve
between inflation and unemployment evolved over time across European
countries, and which structural or policy-related factors are associated
with these changes?**
:::

## Sources

Our two sources are from the [OECD Data Explorer's 117 Economic
Outlook](https://data-explorer.oecd.org/vis?pg=0&snb=1&vw=ov&df%5Bds%5D=dsDisseminateFinalDMZ&df%5Bid%5D=DSD_EO%40DF_EO&df%5Bag%5D=OECD.ECO.MAD&df%5Bvs%5D=&lb=bt&dq=DEU%2BAUT%2BBEL%2BDNK%2BESP%2BEST%2BFIN%2BFRA%2BGRC%2BHUN%2BISL%2BIRL%2BITA%2BLVA%2BLTU%2BLUX%2BNOR%2BNLD%2BPOL%2BPRT%2BSVK%2BGBR%2BSVN%2BSWE%2BCHE%2BBGR%2BHRV%2BROU.IRS%2BIRCB%2BIRL%2BUN%2BES%2BWSST%2BEE%2BECSA%2BEG%2BHRS%2BLF%2BLFPR1574%2BET%2BET_NA%2BPOP%2BULC%2BPOP1574%2BCPI%2BETPT%2BGDPVTR%2BGDPTR%2BET_ANNPCT%2BTTRADE%2BWRT%2BWAGE%2BPCORE_YTYPCT%2BCPI_YTYPCT%2BUNR%2BPDTY.A&pd=%2C&ly%5Brw%5D=REF_AREA&ly%5Bcl%5D=TIME_PERIOD&ly%5Brs%5D=MEASURE&to%5BTIME_PERIOD%5D=false&lc=fr)
and the [World Development
Indicator](https://databank.worldbank.org/source/world-development-indicators#).

::: callout-note
On every data source, we took the maximal range of time available, even
if all data was `NA` and for all European countries. Please note that
European countries does not restrict to solely the European Union, Euro
Area or Euro Zone. It, however, does not include neither Turkey, nor
Russia. As you can see [here](#basic-info), for WDI, we took a
consequent number of variables and therefore you will understand that we
will not display all of them here. You can look inside of the data for
easy replication of the selection.
:::

The two dataset are from major institutions : the OECD and the World
Development indicators :

-   The [OECD](https://www.oecd.org/fr/about.html) is an international
    institution founded in 1961 to advise countries on their public
    policies with the aim of improving the living conditions of their
    citizens. The institution works on a wide range of issues, such as
    education and the fight against global warming. It has been
    collecting and analyzing data for its member countries for 60 years.
    the OECD is composed of 38 countries based on the following
    criteria: a democratic system of government and a market economy. It
    makes available to the public periodic specialised reports, as well
    as ad hoc reports for certain areas, statistical data, country
    studies, and working papers. Thus, the OECD seemed to us to be a
    reliable source for obtaining economic data on the different
    countries of Europe. We selected OECD Data explorer's 117 Economic
    outlook. indeed, The OECD Economic Outlook presents the OECD's
    analysis of the main global economic trends and forecasts for the
    next two years. It offers a coherent set of projections for output,
    employment, government spending, prices and current accounts, based
    on an examination of each member country and the knock-on effect on
    each of them in the context of international developments.

-   The World Development Indicators constitute the principal collection
    of development data compiled by the World Bank. The database brings
    together more than a thousand indicators covering a wide range of
    development topics such as economic growth, poverty, education,
    health, environment, and international trade. The WDI data are drawn
    from officially recognized international sources and are subject to
    strict verification and harmonisation processes to ensure
    comparability across countries and over time. The World Bank has
    been collecting and publishing development statistics for several
    decades as part of its mission to reduce poverty and support
    sustainable development. The WDI therefore represent a reliable and
    widely used source for conducting international economic analyses,
    particularly because they provide consistent, regularly updated, and
    publicly accessible data for most countries in the world.

## Data

::: callout-warning
For simplicity, we continued the "final phase" on the same quarto
document as the "second phase", hence, we did not feel that we needed
another document. Every dataset description and essential variables'
description is in the following chunk.
:::

### Data loading

```{r, message = FALSE, warning = FALSE}
# Loading data
OECD <- vroom("data/DATA OCDE.csv")
WDI <- vroom("data/DATA WDI.csv")
```

### Necessary data manipulation

All the data in OECD is annual and there are many unnecessary columns.
We keep only the reference area, measure, time period and observation
value of the measure, respectively named `country`, `measure`, `time`
and `value`. We will also rename all variables such that they will be
identifiable from the OECD database, but still recognizable.

```{r, message = FALSE, warning = FALSE}
# Manipulating OECD data

OECD <- OECD |>
  select(`Reference area`, Measure, TIME_PERIOD, OBS_VALUE) |>
  rename(
    country = `Reference area`,
    measure = Measure,
    time = TIME_PERIOD,
    value = OBS_VALUE
  ) |>
  pivot_wider(
    names_from = measure,
    values_from = value
  ) |>
  rename(
    nom.eff.ex.rate.oecd = `Nominal effective exchange rate, chain-linked, overall weights`,
    labor.prod.oecd = `Labour productivity, total economy`,
    unit.labor.cost.oecd = `Unit labour cost in total economy`,
    real.eff.ex.rate.oecd = `Real effective exchange rate, constant trade weights`,
    terms.trade.oecd = `Terms of trade, goods and services`,
    ex.rate.LCU.oecd = `Exchange rate, national currency per USD`,
    pot.y.vol.oecd = `Potential output volume`,
    emp.tot.oecd = `Total employment (labour force survey basis)`,
    wages.oecd = `Wages, total economy`,
    compensation.rate.oecd = `Compensation rate, total economy`,
    wage.rate.oecd = `Wage rate, total economy`,
    pot.y.nom.oecd = `Potential output, nominal value`,
    emp.country.spec.oecd = `Employment, country specific`,
    pot.emp.oecd = `Potential employment of total economy`,
    tot.pop.oecd = `Total population`,
    work.age.oecd = `Working-age population, age 15-74`,
    gov.emp.oecd = `General government employment`,
    emp.tot.nat.account.oecd = `Total employment (national accounts basis)`,
    dependent.emp.oecd = `Dependent employment, total economy`,
    self.emp.tot.oecd = `Total self-employed`,
    unemp.level.oecd = `Unemployment, level`,
    labor.force.oecd = `Labour force`,
    cpi.oecd = `Consumer price index`,
    emp.tot.gr.oecd = `Total employment, growth`,
    st.IR.oecd = `Short-term interest rate`,
    lt.IR.gov.bonds.oecd = `Long-term interest rate on government bonds`,
    ecb.IR.oecd = `Central bank key interest rate`,
    labor.force.participation.work.age.oecd = `Labour force participation rate, as a percentage of population aged 15-74`,
    core.infl.oecd = `Core inflation`,
    hours.per.work.oecd = `Hours worked per worker, total economy`,
    headline.infl.oecd = `Headline inflation`,
    unemp.rate.oecd = `Unemployment rate`,
    ex.rate.usd.oecd = `Exchange rate, USD per national currency`
  ) |>
  pivot_longer(
    cols = -c("country","time"),
    names_to = "measure",
    values_to = "value"
  ) |>
  mutate(
    time = as.integer(time),
    value = as.numeric(value)
  ) |>
  drop_na()
```

WDI data is also annual and we also need to get rid of unnecessary
columns. We will also use `pivot_longer` in order to group all the
different columns for the different variables in a new `measure` column
and their values in the `value` one. Also, `Time` will be renamed as
`time`.

```{r, message = FALSE, warning = FALSE}
# Manipulating WDI

WDI <- WDI |>
  select(-c(`Time Code`, `Country Code`)) |>
  rename(
    country = `Country Name`,
    time = Time,
    cpi.wdi = `Inflation, consumer prices (annual %) [FP.CPI.TOTL.ZG]`,
    unemp.youth.tot.wdi = `Unemployment, youth total (% of total labor force ages 15-24) (national estimate) [SL.UEM.1524.NE.ZS]`,
    unemp.tot.wdi = `Unemployment, total (% of total labor force) (national estimate) [SL.UEM.TOTL.NE.ZS]`,
    unemp.male.wdi = `Unemployment, male (% of male labor force) (national estimate) [SL.UEM.TOTL.MA.NE.ZS]`,
    unemp.female.wdi = `Unemployment, female (% of female labor force) (national estimate) [SL.UEM.TOTL.FE.NE.ZS]`,
    unemp.inter.male.wdi = `Unemployment with intermediate education, male (% of male labor force with intermediate education) [SL.UEM.INTM.MA.ZS]`,
    unemp.inter.female.wdi = `Unemployment with intermediate education, female (% of female labor force with intermediate education) [SL.UEM.INTM.FE.ZS]`,
    unemp.inter.tot.wdi = `Unemployment with intermediate education (% of total labor force with intermediate education) [SL.UEM.INTM.ZS]`,
    unemp.basic.male.wdi = `Unemployment with basic education, male (% of male labor force with basic education) [SL.UEM.BASC.MA.ZS]`,
    unemp.basic.female.wdi = `Unemployment with basic education, female (% of female labor force with basic education) [SL.UEM.BASC.FE.ZS]`,
    unemp.basic.tot.wdi = `Unemployment with basic education (% of total labor force with basic education) [SL.UEM.BASC.ZS]`,
    unemp.adv.male.wdi = `Unemployment with advanced education, male (% of male labor force with advanced education) [SL.UEM.ADVN.MA.ZS]`,
    unemp.adv.female.wdi = `Unemployment with advanced education, female (% of female labor force with advanced education) [SL.UEM.ADVN.FE.ZS]`,
    unemp.adv.tot.wdi = `Unemployment with advanced education (% of total labor force with advanced education) [SL.UEM.ADVN.ZS]`,
    cover.benef.ALMP.rich.wdi = `Coverage of unemployment benefits and ALMP in richest quintile (% of population) [per_lm_alllm.cov_q5_tot]`,
    cover.benef.ALMP.poor.wdi = `Coverage of unemployment benefits and ALMP in poorest quintile (% of population) [per_lm_alllm.cov_q1_tot]`,
    cover.benef.ALMP.tot.wdi = `Coverage of unemployment benefits and ALMP (% of population) [per_lm_alllm.cov_pop_tot]`,
    adequacy.cover.ALMP.wdi = `Adequacy of unemployment benefits and ALMP (% of total welfare of beneficiary households) [per_lm_alllm.adq_pop_tot]`,
    benef.results.poor.wdi = `Benefit incidence of unemployment benefits and ALMP to poorest quintile (% of total U/ALMP benefits) [per_lm_alllm.ben_q1_tot]`,
    wage.male.wdi = `Wage and salaried workers, male (% of male employment) (modeled ILO estimate) [SL.EMP.WORK.MA.ZS]`,
    wage.female.wdi = `Wage and salaried workers, female (% of female employment) (modeled ILO estimate) [SL.EMP.WORK.FE.ZS]`,
    wage.tot.wdi = `Wage and salaried workers, total (% of total employment) (modeled ILO estimate) [SL.EMP.WORK.ZS]`,
    deflator.wdi = `Inflation, GDP deflator (annual %) [NY.GDP.DEFL.KD.ZG]`,
    cpi.index.wdi = `Consumer price index (2010 = 100) [FP.CPI.TOTL]`
  ) |>
  pivot_longer(
    cols = -c(time, country),
    names_to = "measure",
    values_to = "value"
  ) |>
  mutate(
    time = as.integer(time),
    value = as.numeric(ifelse(value == "..", NA, value))
  ) |>
  drop_na()
```

### Basic information on our dataset {#basic-info}

```{r}
WDI |>
  summarise(
    "Countries" = n_distinct(country),
    "Variables" = n_distinct(measure),
    "Start date" = min(time),
    "End date" = max(time)
  ) |>
  kable()
```

The WDI dataset has 24 variables for 58 countries and has a time period
from 1960 to 2024.

```{r}
OECD |>
  summarise(
    "Countries" = n_distinct(country),
    "Variables" = n_distinct(measure),
    "Start date" = min(time),
    "End date" = max(time)
  ) |>
  kable()
```

The WDI dataset has 33 variables for 28 countries and has a time period
from 1960 to 2026.

We can now merge both datasets and get some useful statistics out of
them:

```{r, message = FALSE, warning = FALSE}
# Merging data
data_long <- bind_rows(
  OECD |>
    filter(country %in% intersect(OECD$country, WDI$country)),
  WDI |>
    filter(country %in% intersect(OECD$country, WDI$country))
) |>
  drop_na()

# Summarizing data

data_long |>
  summarise(
    "Countries" = n_distinct(country),
    "Variables" = n_distinct(measure),
    "Start date" = min(time),
    "End date" = max(time)
  ) |>
  kable()

# Transforming the data wide for further use
data_wide <- data_long |>
  pivot_wider(
    names_from = "measure",
    values_from = "value"
  )
```

Therefore, our total dataset comprises 28 countries for 57 variables,
and spans 1960 to 2026. Of course, for analysis purposes, we will reduce
this dataset and use subsets, which will be specified every time.

## Focus variables: unemployment, inflation, and wages

Our two main variables are evidently unemployment and inflation, because
they are the variables on which we will focus.

::: callout-tip
## Definitions

The Consumer Price Index (CPI) is a statistical index that tracks the
average change over time in prices paid by households for a fixed
representative “basket” of goods and services.\
The (CPI) inflation rate is the percentage change in CPI over a period
(month-to-month, year-to-year), given by: $$
\pi_t= \frac{CPI_t-CPI_{t-1}}{CPI_{t-1}}\times 100\%
$$ According to OECD definition, an unemployed individual is a working
age individual that is not employed, is available to work and has been
actively looking for employment during a reference period.\
The unemployment rate is therefore defined as the percentage share of
unemployed individuals in total labor force, given by: $$
u_t=\frac{U_t}{L_t}\times 100\%
$$ Where $U_t$ is the total number of unemployed individuals and $L_t$
the total number of individuals in the labor force.

Still according to OECD definition, average annual wages correspond to
the annual rates paid per employee in full-time equivalent units in the
total economy. The wage growth rate is the percentage change in average
annual wages over a period:

$$w_t= \frac{W_t-W_{t-1}}{W_{t-1}}\times 100\%$$ Wages are observed
because the original Phillips curve described the inverse relationship
between unemployment and wage growth.
:::

WDI data will be used because of more completeness for the unemployment
rate and inflation rate variables. Moreover, we exclude inflation values
above 40% as in [Yamada & Bell
(2012)](https://ejce.liuc.it/18242979201202/182429792012090203.pdf). Let
us focus on France:

```{r}
data_wide |>
  filter(country == "France") |>
  drop_na(cpi.wdi) |>
  ggplot(aes(x = time, y = cpi.wdi)) +
  geom_line(na.rm = TRUE) +
  labs(
    title = "Consumer Price Index (France)",
    x = "Year",
    y = "CPI (WDI)"
  ) +
  theme_minimal()
data_wide |>
  filter(country == "France") |>
  drop_na(unemp.tot.wdi) |>
  ggplot(aes(x = time, y = unemp.tot.wdi)) +
  geom_line(na.rm = TRUE) +
  labs(
    title = "Total Unemployment Rate (France)",
    x = "Year",
    y = "Unemployment (WDI, % of labor force)"
  ) +
  theme_minimal()
data_wide |>
  filter(country == "France") |>
  drop_na(wage.rate.oecd) |>
  ggplot(aes(x = time, y = wage.rate.oecd/1000)) +
  geom_line(na.rm = TRUE) +
  labs(
    title = "Average Annual Wages in France (OECD)",
    x = "Year",
    y = "Average Annual Wage (thousands of EUR)"
  ) +
  theme_minimal()
```

## Exogenous variables of interest

The exogenous variables are selected to account for factors that
influence inflation dynamics independently of labor market tightness,
and that may therefore affect the slope and stability of the Phillips
curve. Monetary policy variables, namely the ECB short-term policy rate
and long-term government bond yields, are included to capture the impact
of financial conditions and inflation expectations on price-setting
behavior. By affecting aggregate demand and firms’ financing costs,
these variables can modify the sensitivity of inflation to unemployment.

The exchange rate is introduced to control for external inflationary
pressures arising from import prices and international competitiveness,
which are particularly relevant for open economies. Supply-side factors
are captured through labour productivity and potential output, as these
variables directly influence unit labor costs and production
constraints. Higher productivity growth allows wage increases to be
absorbed without generating inflation, while deviations of output from
its potential level reflect capacity constraints that can amplify or
dampen inflationary pressures. Including these exogenous variables
therefore allows for a more accurate identification of the relationship
between unemployment and inflation underlying the Phillips curve.

### ECB rate

The short-term policy interest rate corresponds to the main refinancing
operations rate set by the European Central Bank (ECB). It represents
the cost at which banks can obtain short-term liquidity from the central
bank and constitutes the primary monetary policy instrument used to
influence aggregate demand and inflation.

```{r}
data_wide |>
  filter(country == "France") |>
  drop_na(ecb.IR.oecd) |>
  ggplot(aes(x = time, y = ecb.IR.oecd)) +
  geom_line(na.rm = TRUE) +
  labs(
    title = "Short term directory rate (ECB)",
    x = "Year",
    y = "ECB rate (%, OECD)"
  ) +
  theme_minimal()
```

### Government bonds' rate

The government bond interest rate refers to the yield on long-term
sovereign bonds, typically measured using 10-year government bonds. It
reflects long-term financing conditions as well as expectations about
future inflation and economic activity.

```{r}
data_wide |>
  filter(country == "France") |>
  drop_na(lt.IR.gov.bonds.oecd) |>
  ggplot(aes(x = time, y = lt.IR.gov.bonds.oecd)) +
  geom_line(na.rm = TRUE) +
  labs(
    title = "10 years government bonds",
    x = "Year",
    y = "Bond rate (%, OECD)"
  ) +
  theme_minimal()
```

### Exchange rate

The exchange rate measures the price of the domestic currency in terms
of a foreign currency, here defined as the bilateral nominal exchange
rate vis-à-vis the US dollar. It is denoted by $e_t$.\
An increase in $e_t$ corresponds to a depreciation of the domestic
currency. Exchange rate movements affect inflation through import prices
and external competitiveness.

```{r}
data_wide |>
  filter(country == "France") |>
  drop_na(ex.rate.usd.oecd) |>
  ggplot(aes(x = time, y = ex.rate.usd.oecd)) +
  geom_line(na.rm = TRUE) +
  labs(
    title = "Exchange rate EUR-USD",
    x = "Year",
    y = "Exchange rate (EUR/USD, OECD)"
  ) +
  theme_minimal()
```

### Labour productivity

Labour productivity is defined as output per unit of labor input and is
measured as real output per employed person. It captures efficiency
gains in the production process and influences firms’ pricing and
wage-setting behavior. Labour productivity is denoted by:
$a_t \equiv \frac{Y_t}{N_t}$ where $Y_t$ denotes real output and $N_t$
the number of employed individuals.

```{r}
data_wide |>
  filter(country == "France") |>
  drop_na(labor.prod.oecd) |>
  ggplot(aes(x = time, y = labor.prod.oecd)) +
  geom_line(na.rm = TRUE) +
  labs(
    title = "Labor productivity",
    x = "Year",
    y = "Labour productivity (OECD)"
  ) +
  theme_minimal()
```

### Potential Output (Volume)

Potential output in volume corresponds to the level of real output that
the economy can sustain in the long run without generating inflationary
pressures. It is determined by structural factors such as technology,
labor supply, and capital accumulation.

```{r}
data_wide |>
  filter(country == "France") |>
  drop_na(pot.y.vol.oecd) |>
  ggplot(aes(x = time, y = pot.y.vol.oecd)) +
  geom_line(na.rm = TRUE) +
  labs(
    title = "Potential Output in Volume",
    x = "Year",
    y = "Potential Output in Volume (OECD)"
  ) +
  theme_minimal()
```

# Introduction to a bit of history

## What is the Phillips curve?

The curve was conceptualized by William Phillips in 1958 in the
Economica journal. He apparently observed the negative relationship
between the increase in nominal wages and unemployment between 1861 and
1957.

::: callout-tip
## Fun fact!

A few people actually know the true story behind this curve, but
according to a friend of Phillips who then worked at the ECB and passed
down this story in the Financial Analysis department, Phillips actually
computed the curve ... by hand! In fact, he took the data, created 5 or
so averages and observed the negative relationship in the graph. This is
the true (yet not so sexy and serious) story behind the "statistical
observation" of this curve. Although his work was thereafter taken by
other brilliant economist, which found (not so strong though) similar
relationships.
:::

He derived the following relationship:

$$
\frac{\Delta W}{W}=\alpha\cdot U^{-\beta}-\gamma
$$

Where $\frac{\Delta W}{W}$ represents the rate of change of nominal
wages and $U$ represents unemployment. The rest are parameters. This
equation has a very practical theoretical reasoning which serves as a
strong foundation for wage bargaining theory, which itself lies in the
center of labor market economics. When unemployment increases, the firm
has more "supply" of potential workers, therefore, their bargaining
power decreases, and so does their wages, which in turns create a
decrease in prices, because firms can supply more. Vice-versa if
unemployment decreases.

## Historical development

In November 1958, the journal Economica published an article by A.W.
Phillips, a New Zealand economist working at the London School of
Economics, entitled ‘The relationship between unemployment and the rate
of growth of nominal wages in the United Kingdom, 1861-1957’. Phillips
sought to test the idea that unemployment and nominal wage growth move
in opposite directions. His intuition was that the labor market is like
any other market: periods of strong demand put upward pressure on
prices, and vice-versa. Thus, when economic activity is strong,
unemployment is low and employees are in a position of strength to
demand wage increases. Conversely, when economic activity slows, high
unemployment encourages wage moderation. The unemployment rate would
therefore be the main determinant of wage growth. An analysis of UK data
confirms the existence of an inverse relationship between the
unemployment rate and the rate of growth in nominal wages.

It enabled Keynesian macroeconomics to move from static to dynamic
analysis, thereby providing the missing equation in Keynesian theory.
Indeed, Keynesian analysis prior to the Phillips curve was limited to
explaining fluctuations in activity through variations in demand, but as
a fixed-price analysis, it shed no light on price dynamics. After the
introduction of the Phillips curve, the macroeconomic equilibrium for a
given period described by the aggregate supply-aggregate demand (AD-AS)
model of synthesis changes over time according to the dynamics of
nominal wages: at fixed prices, at a given period $t$, demand determines
output and employment and therefore also unemployment, which, via the
Phillips relationship, will influence the evolution of nominal wages,
and therefore prices, between period $t$ and period $t + 1$. Prices at
$t + 1$ will in turn determine demand at $t + 1$, and so on.

This relationship, in a later version, had very strong implications for
economic policy. Indeed, an extension of the Phillips curve was proposed
in 1960 by Samuelson and Solow. They first found results similar to
those of Phillips on American data for the period 1900-1960; then, based
on the principle that any change in wages necessarily affects prices via
costs, they transformed Phillips' original relationship into a
relationship between unemployment and inflation (rather than wage
inflation). The Phillips relationship in the form provided by Samuelson
and Solow thus provided evidence of a possible trade-off between
inflation and unemployment.

## Actual researches on it

The newest researches go as far as to develop an open-economy Phillips
curve, where inflation is influenced by both wage levels and the
exchange rate:

$$
\pi_t=\gamma_1 \cdot \mathbb{E}_t[\pi_{t+1}]+\gamma_2 \cdot x_t + \gamma_3 \cdot q_t
$$

Where $\gamma_1$, $\gamma_2$ and $\gamma_3$ are parameters, $\pi_t$
represents domestic inflation, $x_t$ is the (real) output gap and $q_t$
the (real) exchange rate. This equation is based on [Gali & Monacelli
(2005)](https://crei.cat/wp-content/uploads/users/pages/roes8739.pdf).
See [Appendix A](#appendixA) for full derivation.

# Data visualization

Let us represent as an "entrée" a basic scatter plot for the Phillips
curve for 28 European countries over the period and we are using WDI
data because it is more complete. We still exclude inflation rates above
40%.

```{r, message = FALSE, warning = FALSE}
# First page

data_wide |>
  filter(cpi.wdi <= 40) |>
  select(time, country, cpi.wdi, unemp.tot.wdi) |>
  ggplot(aes(x = unemp.tot.wdi, y = cpi.wdi)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "black") +
  facet_wrap_paginate(~ country, ncol = 4, nrow = 3, page = 1, scales = "free") +
  labs(
    title = "Phillips Curve",
    x = "Unemployment rate (%)",
    y = "Inflation rate (%)"
  ) +
  theme_minimal()

# Second page

data_wide |>
  filter(cpi.wdi <= 40) |>
  select(time, country, cpi.wdi, unemp.tot.wdi) |>
  ggplot(aes(x = unemp.tot.wdi, y = cpi.wdi)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "black") +
  facet_wrap_paginate(~ country, ncol = 4, nrow = 3, page = 2, scales = "free") +
  labs(
    x = "Unemployment rate (%)",
    y = "Inflation rate (%)"
  ) +
  theme_minimal()

# Third page

data_wide |>
  filter(cpi.wdi <= 40) |>
  select(time, country, cpi.wdi, unemp.tot.wdi) |>
  filter(country %in% c("Spain", "Sweden", "Switzerland", "United Kingdom")) |>
  ggplot(aes(x = unemp.tot.wdi, y = cpi.wdi)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "black") +
  facet_wrap(~ country, scales = "free") +
  labs(
    x = "Unemployment rate (%)",
    y = "Inflation rate (%)"
  ) +
  theme_minimal()
```

A simple visualization reassures us on theory. We can indeed see that
over the complete span 1960-2026, the Phillips curve seems, for most
countries, to obey the theory and display a downward slopping tendency.
There are though some surprising results, as we can see mainly for the
"Eastern Block" countries, which display almost flat or slightly upward
sloping curves.

Whether these results are due to true economic tendencies or simply a
lack of data or data mismatch will be determined with the analysis.

The question of the curvature of the Phillips curve, however, requires
the differentiation between periods, say decades or key dates, which
will be our next observation.

## Before/after 2000

We focus on 2000 as the dividing year because it corresponds to a major
structural transformation in the European monetary and macroeconomic
framework. The turn of the millennium is closely associated with the
introduction of the euro and the establishment of a single monetary
policy under the European Central Bank, which profoundly reshaped
inflation dynamics and policy transmission mechanisms across euro area
countries. The move from national monetary regimes to a centralized
framework altered inflation expectations, wage-setting behavior, and the
interaction between labor markets and prices.

Moreover, the period around 2000 marks a phase of increased financial
integration and credibility of monetary policy in Europe, accompanied by
a stronger anchoring of inflation expectations. These developments
likely affected the slope and stability of the Phillips curve by
weakening the direct link between domestic labor market conditions and
inflation outcomes. Using 2000 as a breakpoint therefore allows us to
assess how the unemployment–inflation relationship evolved before and
after the adoption of the euro and the consolidation of a common
monetary policy regime.

```{r, message = FALSE, warning = FALSE}
# First page

data_wide |>
  filter(cpi.wdi <= 40) |>
  mutate(
    period = if_else(time < 2000, "Before 2000", "After 2000")
  ) |>
  select(time, country, cpi.wdi, unemp.tot.wdi, period) |>
  ggplot(aes(x = unemp.tot.wdi, y = cpi.wdi, color = period)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  facet_wrap_paginate(~ country, ncol = 4, nrow = 3, page = 1, scales = "free") +
  labs(
    title = "Phillips Curve",
    x = "Unemployment rate (%)",
    y = "Inflation rate (%)"
  ) +
  scale_color_manual(values = c(
  "Before 2000" = "darkgrey",
  "After 2000"  = "black"
)) +
  theme_minimal()

# Second page

data_wide |>
  filter(cpi.wdi <= 40) |>
  mutate(
    period = if_else(time < 2000, "Before 2000", "After 2000")
  ) |>
  select(time, country, cpi.wdi, unemp.tot.wdi, period) |>
  ggplot(aes(x = unemp.tot.wdi, y = cpi.wdi, color = period)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  facet_wrap_paginate(~ country, ncol = 4, nrow = 3, page = 2, scales = "free") +
  labs(
    x = "Unemployment rate (%)",
    y = "Inflation rate (%)"
  ) +
  scale_color_manual(values = c(
  "Before 2000" = "darkgrey",
  "After 2000"  = "black"
)) +
  theme_minimal()

# Third page

data_wide |>
  filter(cpi.wdi <= 40) |>
  mutate(
    period = if_else(time < 2000, "Before 2000", "After 2000")
  ) |>
  select(time, country, cpi.wdi, unemp.tot.wdi, period) |>
  filter(country %in% c("Spain", "Sweden", "Switzerland", "United Kingdom")) |>
  ggplot(aes(x = unemp.tot.wdi, y = cpi.wdi, color = period)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  facet_wrap(~ country, scales = "free") +
  labs(
    x = "Unemployment rate (%)",
    y = "Inflation rate (%)"
  ) +
  scale_color_manual(values = c(
  "Before 2000" = "darkgrey",
  "After 2000"  = "black"
)) +
  theme_minimal()
```

The Phillips Curve graphs compare inflation and unemployment rates
before and after 2000 across various countries. Before 2000, many
nations, like France and Spain, show a clear inverse relationship,
aligning with the traditional Phillips Curve theory. However, post-2000,
this relationship often weakens, disappears or even reverse, as seen in
Sweden or the UK, suggesting shifts in economic dynamics.

Furthermore, some countries, in particular eastern countries, lack
pre-2000 data, which renders comparison complicated and pushes to
period-wise comparison. For that, it is better to either reduce sample
and focus on decade-wise comparison or keep a sample large enough and
compare between large enough periods.

# Data analysis

## Multi-period Difference-in-Difference type model {#did-model}

This model allows us, contrary to a classical Difference-in-Difference
(DiD) model, to analyze the cross-period changes of the Phillips curve
over time, especially the impacts of the underlying variables'
(policies, sensibilities, shocks...) period-specific and cross-period
influences on the slope and the curvature. Furthermore, from [Appendix
A](#appendixA) we know that thanks to Okun's law we can express the NKPC
in terms of economic slack, *i.e.*, unemployment gap.

Formally, let $\pi_t$ denote inflation, $u_t$ unemployment gap, then
$u_t^2$ captures the nonlinear sensitivity to slack, $D_{k,t}$ denotes
the $k$-th period dummy variable, which takes values in $\{0,1\}$,
indexed by $k=1,...,K-1$ and $Z_{j,t}$ the underlying structural or
policy variables, indexed by $j=1,...,J$.

The model specification is:

$$
\begin{aligned}
\pi_t 
= &\;
\underbrace{\alpha}_{\text{Intercept}} +
\underbrace{\beta_1 u_t}_{\text{Baseline slope}}
+
\underbrace{\beta_2 u_t^{2}}_{\text{Baseline curvature}}
\\
&+
\underbrace{\sum_{k=1}^{K-1} \delta_k D_{k,t}}_{\text{Period fixed effects}}
+
\underbrace{\sum_{j=1}^{J} \gamma_j Z_{j,t}}_{\text{Direct effects of underlying variables}}
\\
&+
\underbrace{\sum_{k=1}^{K-1} \theta_{1k} (u_t \times D_{k,t})}_{\text{Slope differences across periods}} +
\underbrace{\sum_{k=1}^{K-1} \theta_{2k} (u_t^{2} \times D_{k,t})}_{\text{Curvature differences across periods}} \\ &+
\underbrace{\sum_{j=1}^{J} \phi_{1j} (u_t \times Z_{j,t})}_{\text{Effect of underlying variables on slope}} + \underbrace{\sum_{j=1}^{J} \phi_{2j} (u_t^{2}\times Z_{j,t})}_{\text{Effect of underlying variables on curvature}} \\ &+
\underbrace{
\sum_{k=1}^{K-1}\sum_{j=1}^{J} 
\psi_{1kj} (u_t \times Z_{j,t} \times  D_{k,t})
}_{\text{Period-specific slope effects of underlying variables}}
\\ &+
\underbrace{
\sum_{k=1}^{K-1}\sum_{j=1}^{J} 
\psi_{2kj} (u_t^{2} \times Z_{j,t} \times D_{k,t})
}_{\text{Period-specific curvature effects of underlying variables}} \\ &+ \varepsilon_t.
\end{aligned}
$$ Where $\beta_1$ and $\beta_2$ specify the baseline of the model,
indicating the slope sign and the curvature, or sensitivity, of the
slope, respectively, $\theta_{1k}$ and $\theta_{2k}$ are period
effects-specific parameters and indicate how the slope and the curvature
are respectively affected, $\phi_{1j}$ and $\phi_{2j}$ are underlying
effects-specific parameters and indicate how the slope and the curvature
are respectively affected, and $\psi_{1kj}$ and $\psi_{2kj}$ are
cross-period, underlying effects-specific parameters indicating the
respective differing effects (if there are) on the slope and the
curvature of the curve.

::: callout-note
## Note

Contrary to the usual binary DiD model, which is often found in the
form: $$
Y_{it} = \alpha + \beta \, (Treat_i \times Post_t) + \gamma_i + \delta_t + \varepsilon_{it}
$$ And which concerns the differences between a "treatment" and a
"control" group, this experimental model concerns a more generalized
approach, including non-linearity, more-than-two interaction terms and
cross-period parameters. Inspiring sources using this kind of method in
economics are niche, but not inexistant, seeing [Baker et al.
(2025)](https://arxiv.org/pdf/2503.13323), [Callaway & Sant'Anna
(2018)](https://www.cla.temple.edu/RePEc/documents/DETU_18_04.pdf),
[(2020)](https://psantanna.com/files/Callaway_SantAnna_2020.pdf), or
[Chaisemartin et al. (2022)](https://arxiv.org/pdf/2201.06898).
:::

## Model implementation

```{r}
Z_vars <- c(
  "ecb.IR.oecd",
  "lt.IR.gov.bonds.oecd",
  "ex.rate.usd.oecd",
  "labor.prod.oecd",
  "pot.y.vol.oecd"
)

panel_df <- data_wide |>
  arrange(country, time) |>
  group_by(country) |>
  mutate(
    # Inflation: only compute log-diff when CPI > 0
    pi = 100 * (cpi.wdi - lag(cpi.wdi))/lag(cpi.wdi),

    # Unemployment gap (within-country)
    u_gap = unemp.tot.wdi - mean(unemp.tot.wdi, na.rm = TRUE),

    # Periods
    period = case_when(
      time < 1999 ~ "pre_euro",
      time < 2008 ~ "euro_pre_gfc",
      time < 2020 ~ "post_gfc",
      TRUE        ~ "post_covid"
    )
  ) |>
  ungroup() |>
  select(country, time, pi, u_gap, period, all_of(Z_vars)) |>
  drop_na()
```

```{r}
df <- panel_df |>
  mutate(
    u2 = u_gap^2,
    period = factor(period),
    country = factor(country)
  )

Z_block <- paste(Z_vars, collapse = " + ")

formula_did <- as.formula(paste0(
  "pi ~ u_gap + u2",                  # β1, β2
  " + period",                        # Σ δ_k D_k,t
  " + country",                      # country FE
  " + ", Z_block,                    # Σ γ_j Z_j,t
  " + u_gap:period + u2:period",     # θ terms
  " + u_gap:(",  Z_block, ")",
  " + u2:(",     Z_block, ")",       # φ terms
  " + u_gap:(",  Z_block, "):period",
  " + u2:(",     Z_block, "):period" # ψ terms
))

fit <- lm(formula_did, data = df)
```

```{r}
vcov_country <- vcovCL(
  fit,
  cluster = ~ country,
  type = "HC1"
)

did_results <- coeftest(fit, vcov = vcov_country)
```

```{r}
coefs <- coef(fit)
periods <- levels(df$period)

marginal_effects <- expand_grid(
  variable = Z_vars,
  period   = periods
) |>
  mutate(
    slope_effect = map2_dbl(variable, period, \(v, p) {
      phi <- coefs[paste0("u_gap:", v)]
      psi <- coefs[paste0("u_gap:", v, ":period", p)]
      phi + ifelse(is.na(psi), 0, psi)
    }),
    curvature_effect = map2_dbl(variable, period, \(v, p) {
      phi <- coefs[paste0("u2:", v)]
      psi <- coefs[paste0("u2:", v, ":period", p)]
      2 * (phi + ifelse(is.na(psi), 0, psi))
    })
  )
```

```{r}
Z_means <- df |>
  group_by(country, period) |>
  summarise(
    across(all_of(Z_vars), \(x) mean(x, na.rm = TRUE)),
    .groups = "drop"
  )
effects_country_period <- Z_means |>
  pivot_longer(
    cols = all_of(Z_vars),
    names_to = "variable",
    values_to = "Zbar"
  ) |>
  left_join(marginal_effects, by = c("variable", "period")) |>
  mutate(
    slope_contribution     = slope_effect * Zbar,
    curvature_contribution = curvature_effect * Zbar
  )
```

```{r}
final_table <- effects_country_period |>
  select(
    country,
    period,
    variable,
    slope_contribution,
    curvature_contribution
  ) |>
  arrange(period, country, desc(abs(slope_contribution)))
ranking_table <- final_table |>
  group_by(period, variable) |>
  summarise(
    avg_abs_slope     = mean(abs(slope_contribution), na.rm = TRUE),
    avg_abs_curvature = mean(abs(curvature_contribution), na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(period) |>
  mutate(
    pct_slope     = 100 * avg_abs_slope / sum(avg_abs_slope),
    pct_curvature = 100 * avg_abs_curvature / sum(avg_abs_curvature)
  ) |>
  ungroup() |>
  arrange(period, desc(pct_slope))

```

```{r}
period_labels <- c(
  pre_euro      = "Pre-EMU (pre-1999)",
  euro_pre_gfc  = "EMU – pre-GFC",
  post_gfc      = "Post-GFC",
  post_covid    = "Post-COVID"
)

var_labels <- c(
  "ecb.IR.oecd"          = "ECB policy rate",
  "lt.IR.gov.bonds.oecd" = "Long-term gov. yield",
  "ex.rate.usd.oecd"    = "USD exchange rate",
  "labor.prod.oecd"    = "Labour productivity",
  "pot.y.vol.oecd"      = "Potential output volatility"
)
```

```{r}
ranking_table |>
  mutate(
    period   = recode(period, !!!period_labels),
    variable = recode(variable, !!!var_labels)
  ) |>
  select(period, variable, pct_slope, pct_curvature) |>
  mutate(
    pct_slope     = round(pct_slope, 1),
    pct_curvature = round(pct_curvature, 1)
  ) |>
  knitr::kable(
    format = ifelse(knitr::is_latex_output(), "latex", "html"),
    booktabs = TRUE,
    caption = "Percentage contribution of structural variables to the Phillips curve, by period",
    col.names = c(
      "Period",
      "Variable",
      "% contribution to slope",
      "% contribution to curvature"
    )
  ) |>
  kableExtra::kable_styling(
    latex_options = c("hold_position", "scale_down"), 
    full_width = FALSE,
    position = "center"
  )
```

### Graphs on contribution

```{r}
ranking_table |>
  mutate(
    period   = recode(period, !!!period_labels),
    variable = recode(variable, !!!var_labels)
  ) |>
  select(period, variable, pct_slope) |>
  ggplot(aes(x = period, y = pct_slope, fill = variable)) +
  geom_col(width = 0.7) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "What shapes the Phillips curve slope? (percentage contributions)",
    x = "",
    y = "% contribution",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(hjust = 0.5)
  )

ranking_table |>
  mutate(
    period   = recode(period, !!!period_labels),
    variable = recode(variable, !!!var_labels)
  ) |>
  select(period, variable, pct_curvature) |>
  ggplot(aes(x = period, y = pct_curvature, fill = variable)) +
  geom_col(width = 0.7) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "What shapes the Phillips curve curvature? (percentage contributions)",
    x = "",
    y = "% contribution",
    fill = ""
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

A detailed country-by-country per period contribution table can be found
in [Appendix B](#appendixB)

# Conclusion

The results show that the slope of the Phillips curve is mainly
determined, regardless of the period, by labor productivity, with an
average contribution of 69.2% over the periods, ranging from 58.1% to
75.9%. This represents the largest share of the explanatory
contribution, suggesting that the transmission of labor market tensions
to inflation occurs primarily through unit labor costs and the ability
of firms to absorb wage increases. Before the 2008 global financial
crisis, particularly in the pre-euro and pre-GFC euro periods, financial
variables such as long-term interest rates and, to a lesser extent,
exchange rates still played a significant role in shaping the slope.
This reflects a macroeconomic environment in which financial conditions
and inflation expectations were less well anchored. Interest rates on
government bonds contributed up to 25% to the slope. After the GFC,
however, the contribution of financial variables declined sharply, while
that of productivity increased, consistent with the idea of a weakening
of the traditional monetary transmission channel and a flattening of the
Phillips curve.\
The post-financial crisis period is thus characterized by an
inflation-unemployment relationship that is driven more by structural
factors than cyclical ones. The weak contribution of ECB policy rates
and long-term rates to the slope of the curve suggests that, in a
context of deleveraging, liquidity traps, and low inflation, monetary
policy has less of a direct influence on price dynamics via the labor
market. This pattern has persisted and intensified since the Covid-19
crisis, despite a more unstable macroeconomic environment, confirming
that the slope of the curve depends primarily on real mechanisms linked
to productivity and the organization of the labor market. Nevertheless,
the relatively stable and low contributions of the exchange rate,
fluctuating between 0.9% and 1.8%, indicate the weak impact of
international labor mobility on the relationship between endogenous
variables.\
With regard to the curvature of the Phillips curve, the results indicate
that it is mainly shaped by labor productivity and potential GDP
volatility, with an average contribution of 60.5% for the former and
24.3% for the latter. The curvature captures nonlinear effects, i.e.,
the fact that inflation reacts more strongly when the economy is
operating close to or above its potential. Before the 2008 crisis,
macroeconomic volatility played an important role in the emergence of
these nonlinearities, reflecting a less stable inflation regime. After
the GFC, the contribution of financial variables to the curvature became
marginal, i.e., never exceeding 10% individually, suggesting that
extreme inflationary pressures no longer stem primarily from financial
conditions, but rather from real supply constraints.\
Finally, the post-Covid period is characterized by a relative increase
in the volatility of potential GDP in explaining the curve, linked to
supply shocks, disruptions in value chains, and sectoral reallocations.
This indicates that, even though the Phillips curve has been flatter
overall since the 2008 financial crisis, it remains nonlinear when the
economy faces major structural shocks. Overall, these results suggest
that the GFC represents a lasting break in the functioning of the
Phillips curve, which has become less sensitive to monetary instruments
and more governed by structural and supply factors.\
The results of this analysis are consistent with the theory of economic
convergence: as economies grow, output converges toward its potential
level and endogenous variables are determined dichotomously by supply.
The Phillips curve relationship therefore exhibits characteristics such
that, during major non-structural shocks, we observe significant
contributions from exogenous variables linked to the stabilization
instruments available to economic institutions and, conversely, a weaker
contribution during periods of stabilization. This naturally leads to
the establishment, in the long term, of economies where a quasi-infinite
number of unemployment levels are compatible with a (convergent) level
of inflation.

# Appendix

## Appendix A: Derivation of the open economy New Keynesian Phillips Curve (NKPC) {#appendixA}

In the classical New Keynesian model, the demand side of the economy
yields the CES consumption function:

$$
C_t(i)=\left(\frac{P_t(i)}{P_t} \right)^{-\varepsilon}\cdot C_t
$$

For $i\in[0,1]$ firms. Market clears at $Y_t(i)=C_t(i)$. Firms produce
according to a linear production function $Y_t(i)=A_tN_t(i)$ where
$a_t\equiv \log{A_t}$ follows an AR(1) process with MA($\infty$)
representation $a_t=\sum_{k=0}^{+\infty}{\rho_a^k \epsilon_{t-k}}$.
Moreover, firms price [*a la Calvo*
(1983)](https://www.sciencedirect.com/science/article/abs/pii/0304393283900600),
where $1-\theta$ (oligopolist) firms can reset prices at level
$P_t^*(i)$ in each period, which implies maximizing:

$$
\max_{P_t(i)} {\mathbb{E}_t \left[\sum_{k=0}^{+\infty}{(\beta\theta)^k \Lambda_{t,t+k} Y_{t+k|t}\left(P_t^*(i)-MC_{t+k}P_{t+k} \right)} \right]}
$$ Under the demand constraint
$Y_{t+k|t}=\left(\frac{P_t^*(i)}{P_t} \right)^{-\varepsilon}\cdot Y_{t+k}$
and the marginal cost of the firm defined as:

$$
MC_t=(1-\chi) \frac{W_t}{P_t}+\chi S_t P_t^f
$$

Where $S_t$ is the nominal exchange rate and $P_t^f$ is the foreign
price index. $\chi$ represents home bias preference, and therefore is a
natural index for openness to trade The optimal consumer price index
(CPI) inflation condition, where inflation is defined as
$\pi_t \equiv p_t-p_{t-1}$, is given by:

$$
\pi_t=\beta \cdot \mathbb{E}_t[\pi_{t+1}]+\kappa \cdot \widehat{mc}_t
$$

Where $\kappa \equiv \frac{(1-\theta)(1-\beta\theta)}{\theta}$. Real
marginal cost $\widehat{mc}_t$ can be found by log-linearizing the
firms' nominal marginal cost, which holds for every firm due to the
linearity of the production function:

$$
\widehat{mc}_t =(1-\chi)\cdot (\hat{w}_t-\hat{p}_t-\hat{a}_t)+\chi\cdot (\hat{s}_t+\hat{p}^f_t-\hat{p}_t) \\
\Leftrightarrow \widehat{mc}^*_t =(\hat{w}_t-\hat{p}_t-\hat{a}_t)+\alpha \cdot \hat{q}_t
$$ Where $\hat{s}_t+\hat{p}^f_t-\hat{p}_t\equiv \hat{q}_t$ is exactly
the formula for the real exchange rate,
$\widehat{mc}^*_t \equiv \frac{\widehat{mc}_t}{1-\chi}$ and
$\alpha \equiv \frac{\chi}{1-\chi}$. This allows to derive the final New
Keynesian Phillips Curve (NKPC):

$$
\pi_t =\beta \cdot \mathbb{E}_t[\pi_{t+1}]+\kappa \cdot (\hat{w}_t-\hat{p}_t-\hat{a}_t)+\kappa\alpha \cdot \hat{q}_t\\
$$ Which allow us to derive three different possible NKPC:

$$
\text{NKPC (1):} \quad \pi_t =\gamma_1 \cdot \mathbb{E}_t[\pi_{t+1}]+\gamma_2 \cdot (\hat{w}_t-\hat{p}_t-\hat{a}_t)+\gamma_3 \cdot \hat{q}_t
$$ With parameters $\gamma_1\equiv\beta$, $\gamma_2\equiv \kappa$ and
$\gamma_3\equiv\kappa \alpha$.

$$
\text{NKPC (2):} \quad \pi_t =\gamma_1 \cdot \mathbb{E}_t[\pi_{t+1}]+\gamma_2 \cdot (\hat{y}_t-\hat{y}^n_t)+\gamma_3 \cdot \hat{q}_t
$$ Using the fact that under imperfect competition, the real unit labor
cost is equal to the output gap such that
$\hat{w}_t-\hat{p}_t-\hat{a}_t=\hat{y}_t-\hat{y}^n_t$, and with
parameters $\gamma_1\equiv\beta$, $\gamma_2\equiv \kappa\varphi$ and
$\gamma_3\equiv\kappa \alpha$.

$$
\text{NKPC (3):} \quad \pi_t =\gamma_1 \cdot \mathbb{E}_t[\pi_{t+1}]+\gamma_2 \cdot (\hat{u}_t-\hat{u}^n_t)+\gamma_3 \cdot \hat{q}_t
$$ Using Okun's law, which can be expressed as
$-\delta(\hat{y}_t-\hat{y}^n_t)=\hat{u}_t-\hat{u}^n_t$, and with
parameters $\gamma_1\equiv\beta$,
$\gamma_2\equiv -\frac{\kappa}{\delta}$ and
$\gamma_3\equiv\kappa \alpha$.

## Appendix B: Results per country {#appendixB}

Here is the country-by-country per period contributions of each
exogenous variables to the slope and curvature of the Phillips curve:

```{r results="asis"}
# ---- detect country column safely ----
country_col <- intersect(
  c("country", "Country", "iso", "iso3", "ctry"),
  names(final_table)
)[1]

if (is.na(country_col)) {
  stop("No country column found in final_table")
}

# ---- prepare data ----
tbl <-
  final_table |>
  group_by(across(all_of(country_col)), period) |>
  mutate(
    pct_slope = 100 * abs(slope_contribution) /
      sum(abs(slope_contribution), na.rm = TRUE),
    pct_curvature = 100 * abs(curvature_contribution) /
      sum(abs(curvature_contribution), na.rm = TRUE)
  ) |>
  ungroup() |>
  select(
    all_of(country_col),
    period,
    variable,
    pct_slope,
    pct_curvature
  ) |>
  rename(country = all_of(country_col)) |>
  mutate(
    pct_slope = round(pct_slope, 1),
    pct_curvature = round(pct_curvature, 1)
  ) |>
  arrange(country, period, desc(pct_slope))

# ---- one table per country ----
for (ctry in unique(tbl$country)) {

  df <- tbl |> filter(country == ctry)

  out <-
    df |>
    select(-country) |>
    knitr::kable(
      format = ifelse(knitr::is_latex_output(), "latex", "html"),
      booktabs = TRUE,
      longtable = knitr::is_latex_output(),
      caption = paste0("Percentage contributions to the Phillips curve — ", ctry),
      col.names = c(
        "Period",
        "Variable",
        "% contribution to slope",
        "% contribution to curvature"
      )
    ) |>
    kableExtra::kable_styling(
      latex_options = if (knitr::is_latex_output()) "repeat_header" else NULL,
      bootstrap_options = c("striped", "condensed"),
      full_width = FALSE,
      position = "center"
    )

  print(out)

  if (!knitr::is_latex_output()) cat("\n\n")
}
```
